---
title: "Titanic Wikipedia Data Grab"
author: "Gina Reynolds"
date: "12/11/2017"
output: pdf_document
---

This file is written to collect the information about those on board Titanic from the Wikipedia pages on passengers and crew. 

```{r}
library(htmltab)
library(dplyr)
library(stringr)
```
```{r}
sessionInfo()
```

# Raw Data
```{r}
if(!dir.exists("RawData")){dir.create("RawData")}

if(!file.exists("RawData/Passengers2017-12-17.html")){
download.file("https://en.wikipedia.org/wiki/Passengers_of_the_RMS_Titanic",
              destfile=paste0("RawData/Passengers",Sys.Date(),".html"))
              }

if(!file.exists("RawData/Crew2017-12-17.html")){
download.file("https://en.wikipedia.org/wiki/Crew_of_the_RMS_Titanic",
              destfile=paste0("RawData/Crew",Sys.Date(),".html"))
              }
```

# Passengers
```{r}
url="RawData/Passengers2017-12-17.html"
Table1=htmltab(url, 1,rm_nodata_cols = F)
Table2=htmltab(url, 2,rm_nodata_cols = F)
Table3=htmltab(url, 3,rm_nodata_cols = F)
Table1$Class="First"
Table2$Class="Second"
Table3$Class="Third"
Passengers=bind_rows(Table1,Table2,Table3);dim(Passengers)
```

# note wikipedia mistake for passengers for Everett, Washington, USA
```{r}
######## passengers #######
Passengers[str_detect(Passengers$Boarded, "Everett"),]
Passengers[str_detect(Passengers$Boarded, "Everett"),"Lifeboat"]=14
Passengers[str_detect(Passengers$Boarded, "Everett"),"Destination"]="Everett, Washington, USA"
Passengers[str_detect(Passengers$Boarded, "Everett"),"Boarded"]=NA
Passengers[c(1025,1026),]
```


# Survival is ID'd with Color... html is style in <tr field.
```{r}
Lines=readLines(url)
BeforeTablesLine=which(str_detect(Lines,'<th>Lifeboat'))
GrabWhich=which(c(rep(T, nrow(Table1)), F, rep(T, nrow(Table2)), F, rep(T, nrow(Table3))))
TempLines=Lines[BeforeTablesLine[1]:length(Lines)]
Passengers$survivaloutcome=str_detect(TempLines[str_detect(TempLines, "<tr")], "style")[GrabWhich]
```

# Crew
```{r}
##### crew ########
url="RawData/Crew2017-12-17.html"
Lines=readLines(url)
BeforeTablesLine=which(str_detect(Lines,'<th>Hometown'))
Crew=data_frame()
for (i in 1:8){
temp=htmltab(url, i,rm_nodata_cols = F)
TempLines=Lines[BeforeTablesLine[i]:length(Lines)]
temp$survivaloutcome=str_detect(TempLines[str_detect(TempLines, "<tr")], "style")[1:nrow(temp)]
Crew=bind_rows(Crew, temp)
}

Crew$Crew="Crew"
Table=bind_rows(Passengers,Crew); dim(Table) 
```

# Join Passenger and Crew Tables
```{r}
# Preparation for full join - some people classified as crew and passengers!
Passengers$Hometown[Passengers$Hometown=="Belfast, Ireland, UK"]="Belfast, Ireland"
Passengers$Name[Passengers$Name=="Frost, Mr. Anthony Wood \"Archie\""]="Frost, Mr. Anthony Wood"
Passengers$Name[Passengers$Name=="Frost, Mr. Anthony Wood \"Artie\""]="Frost, Mr. Anthony Wood"

Table=full_join(Passengers,Crew); dim(Table) # 
Table$Crew[is.na(Table$Crew)]="Not Crew"
Table$survivaloutcome=ifelse(Table$survivaloutcome, "Survived", "Perished")
```




# Sex and Age
```{r}
# Sex
Table$sex=NA
# I inspected titles to see is first names were all male.  There is a Dr. Alice.  
# I overwrite the case below, designating this individual as female.  
# Also, any last names like John, Wallace and the like will be overwriten if there is a woman's title.  
Table$sex[str_detect(Table$Name, "Master |Mr. |Mr |Father |Dr. |Sir |Don |Commander |Captain |Major |Colonel |Reverend |Lieutenant |John |Walter|Georges |Wallace|Theodore |Roger |Percy|George ")]="Male"
Table$sex[str_detect(Table$Name, "Miss |Mrs.|Doña |Countess |Lady |Alice")]="Female"
table(Table$sex, as.numeric(Table$Age)>=18, useNA = "ifany")
table(Table$survivaloutcome,Table$Lifeboat)
Table[is.na(Table$sex),] # These are probably men too - Position Trimmer and Fireman/Stoker

```



# Age
```{r}
# Age
Table$AgeCharacter=Table$Age
table(Table$AgeCharacter, useNA = "ifany")
Table$Age[str_detect(Table$Age,"m")]=0
Table$Age=as.numeric(Table$Age)
table(Table$Age, useNA = "ifany")
```

# Variable Names to Lower Case
```{r}
names(Table)=str_replace(tolower(names(Table))," ","")
```


# Save Data
```{r}
if(!dir.exists("DataProducts")){dir.create("DataProducts")}
str(Table)
save(Table, file = "DataProducts/PeopleOnTitantic.RData")
write.csv(Table, "DataProducts/PeopleOnTitantic.csv", row.names = F)
```

